Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-02-27 paper1_jiturra.Rmd
up

The link between social class and political attitudes constitutes one of the most relevant findings in sociology. Theoretically, studies in the social mobility field have defined classes as structural positions given their difference in labor market situations distinguished by employment relations, skills level, and authority within the workplace, which traditionally has been represented by occupations instead of relative income or educational credentials [@Erikson1992;@Erikson1979]. Besides the social stratification studies, political sociologists have systematically demonstrated that classes trace distinctions in other domains of social life as well, including ideological preferences [@evans_decline_1993], cultural consumption behavior [@Chan2007] and attitudes toward redistribution and welfare [@Svallfors2006]. Regarding the latter, during the last decades, the research on class and political attitudes has developed more extensively regarding the economic domain, addressing the economic interests of classes with an emphasis on redistribution, social security, and social service functions of the welfare state [@lindh_class_2020, p. 421].

Critical views have suggested that economic development and modernization have eroded the link between social class and political attitudes. Around the debate of the "death of class" thesis, it has been argued that due to the increase of material welfare in post-industrial societies, the distributive struggle between classes has been waning, resulting in the salience of other group identities as more relevant, reorienting the political debate towards post-material concerns [@clark_declining_1993;@inglehart_culture_1990]. However, although modernization has led to changes in citizens' concerns, it is feasible to argue that the distributive struggle between classes has not been wholly displaced but cohabits with people's attitudes in the sociocultural domain [@lindh_class_2020]. Empirically, the claimed blurring of the class cleavage has not been strongly supported, whereas class divides indeed shape attitudes in the economic domain in general, such as perceived class conflicts [@perez_class_2023;@edlund_democratic_2015], attitudes toward market-based inequality [@andersen_preferences_2018;@lindh_public_2015], and redistributive preferences in particular [@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010]. In the broader context, scholars have identified two competing mechanisms to elucidate the relationship between social class and attitudes towards inequality, commonly referred to as the self-interest and normative value-driven motives perspectives.

Theoretically, self-interest has served as the canonical model to explain redistributive preferences, where the material interests of those with the most to gain from redistribution conflict with those in positions of greater material well-being [@meltzer_rational_1981]. Similarly, extensions of the model have paid more attention to occupation as a representation of potential unemployment risk, arguing that social classes with greater risk exposure are prone to support redistributive policies as they reduce labor market insecurity [@rehm_risks_2009]. Differently, part of the scholarly discussion has stressed the role of cognitive processes in preference formation for understanding why, even in contexts of greater economic disparities, the lower classes oppose redistributive policies, arguing that misinformation plays a crucial role in linking material conditions with the redistributive policies [@bartels_homer_2005;@druckman_preference_2000]. Similarly, it has been argued that contexts of greater segregation affect the inferences about social and economic inequality [@mijs_inequality_2018] that ultimately contribute to the decoupling of material interests from economic preferences [@hvidberg_social_2020;@garcia-castro_changing_2022].

As material interests might prevail in scarcity, normative or value-driven motivations emerge as an alternative explanation in preference formation when pure self-interest weakens. Thus, values such as self-transcendence or humanitarianism share a common spirit regarding their desire for equality and concern for human dignity, providing the common ground for altruism that ultimately motivates political preferences [@Kulin2013;@feldman_humanitarian_2001]. Despite the protagonism of values, these are indeed conditioned to risk exposure and resource availability, becoming salient under conditions of greater security against market risks and material prosperity and weakening under conditions of greater vulnerability and material precariousness [@Maldonadoetal2019]. Besides, it has been argued that occupations play a role in political attitude formation as they represent a set of specific skills and work logic that convey particular normative views [@oesch_redrawing_2006]. Thus, workspaces are conceived as socializing agents, considering that economically active people spend a substantial part of their time at work, including the social relations involved in providing goods and services. Hence, empirical evidence has suggested that technical and managerial professionals hold more conservative political views than occupations in interpersonal services [@oesch_electoral_2018;@lindh_class_2020]. Despite this, it is challenging to disentangle if the class differences are indeed explained through work experiences because it is feasible to argue that political preferences are the result of early-life internalization of social norms, which might be crystalized in career self-selection paths, which also might be influenced by the organizational hiring requirements for specific jobs [@kitschelt_occupations_2014].

@@

As social relations within the occupational structure serve as the foundation for class analysis, widening our examination to other dimensions of social life, such as family dynamics, the development of friendships, and diverse social activities, can significantly enrich the scope of class analysis, particularly concerning its relevance in political attitudes formation. In principle, social stratification scholars have traditionally paid attention to _differentiation_ as the observable patterns of association given by the intersection of status and group membership, which can be understood as relational networks between positions [@blau_inequality_1977]. Empirically, homophily is one of the most significant findings in network research, whereas the likelihood of tie formation between individuals of similar characteristics is strong [@mcpherson_birds_2001; @lazarsfeld_friendship_1954] around demographic characteristics that trace friendship and family ties such as age and gender, as well as the segregated social environments alongside ethnicity and socioeconomic status [@bargsted_pautas_2020; @plaza_juntos_2022]. In addition, psychosocial factors such as socialization preferences and group dynamics also play a role in segregation processes, considering that isolated social environments decrease the interchange of world views, reinforcing in-group selectivity and value similarity [@visser_attitudes_2004;@homans_human_1951]. Nonetheless, without denying the relevance of socialization preferences, segregation has been predominantly explained through the lens of social differentiation in social activities as foci of contact opportunities that ultimately consolidate social networks [@feld_focused_1981; @mcpherson_network_2019]. From this scope, we argue that class relations not only represent resources-based distinctions, but patterns of sociability and exchange observed in the differentiation of social ties.

How classes are formed also provides insights regarding their social relations and sociability practices. In sociology, classes are understood as the everyday basis for action, given their labor market situation, that ultimately shapes their access to opportunities and resources [@weber_economy_1978]. According to habitus theory, classes incur a series of distinctive practices pivoted in cultural capital as the basis for mutual recognition [@bourdieu_distinction_1984]. Thus, classes seek to improve their position through the intergenerational transmission of resources, whereas similarities in friendship and family choices play a crucial role in the reproduction privilege. Additionally, while symbolic resources reinforce status distinctions, access to exclusive social activities in the upper class increases segregation at the expense of excluding the lower classes [@bourdieu_reproduccion_1981]. Ultimately, classes consolidate their demographic and cultural identities by mobilizing resources and shared sociability practices, ensuring intergenerational reproduction of their structural positions [@goldthorpe_sobre_1992]. Hence, the socialization of common lifestyles and shared worldviews facilitates the consolidation of social classes, whereas social resources embedded in social networks follow a pattern of accumulation and exclusion that ultimately leads to social segregation in both the upper and the lower bounds of the social structure.

Previous research on social class and tie formation has suggested that class permeability, understood as the formation of cross-class social ties, is traced differently across property and authority boundaries. Empirically, evidence from Western industrialized societies has shown that the formation of friendship ties between owners and the working class is very less likely, in contrast to the tendency of higher tie formation between supervisors and workers, suggesting that the authority dimension is much less permeable than authority-based boundaries given that class interests increase the social distance between proprietors and workers, while the intermediate class position of supervisors and contact frequency with workers make friendship tie formation more likely [@wright_relative_1992]. Similarly, evidence from Chile has shown that the higher permeability of the intermediate classes contrasts with the more homogeneous networks of the working class, suggesting that their limited resources and lower capacity to be socially engaged ultimately result in a lack of social resources that isolate them from others, while the upper class is less permeable and homogenous as a result of its tendency to self-selection as a practice that ultimately seeks to reproduce their privileged status positions [@otero_open_2021].

In contrast, cross-class embeddedness can be described regarding their social activities as well as through their changes across generations and time. In this regard, @pichler_social_2009 suggest that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the working class, which tends to be more homogeneous in its civic engagement behavior. Similarly, other studies in Europe and South America have exhibited that the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes, where being socially mobile does help to improve both dimensions in tie formation, whereas it does not equalize the weight of class background compared to those intergenerationally stable in the upper class [@cepic_how_2020;@carrascosa_class_2023]. Additionally, the few longitudinal evidence suggests that networks change following a dynamic of cumulative advantages regarding its composition in terms of prestige and diversity, where upper classes improve in both dimensions while the lower class shows more stability across the life course [@volker_social_2020].

Theoretically, these studies account for class-based network segregation as the degree of connectedness of an individual to different occupations is meant to represent social resources embedded in social networks vertically in the social structure [@lin_social_2007]. Drawing on this approach, the attempts of stratification scholars have been focused on how social connections are distributed across the class structure as a matter of social integration [@blau_inequality_1977]. Besides, _homogeneity_  is described as the lack of cross-class network ties. It is indeed conceptually more proximate to the homophily principle because it is anchored in ego’s class position. At the same time, _diversity_ is defined as the rate of dissimilar ties or simply the total ties to certain groups that do not necessarily count with a reference position to describe the network composition. Hence, both approaches provide a set of alternatives for the study of social class from a network perspective that has been increasingly discussed in the stratification literature as well as in the scholarly discussion about social class and political attitudes.

@@

Differently, it has been suggested that the lens of a network approach provides a better picture of the class relations that nurture social norms and group identity [@kalmijn_social_2007, p. 550]. This claim resembles the fact that classes are characterized as collectivities with differences in their degrees of cohesion and solidarity, encompassing unequal status-based social interactions that are linked to the individual or household material well-being, cultural perspectives, and political preferences that structure the broader social experiences [@morris_attenuation_1996, p. 48]. Furthermore, these social relations contribute to the degree to which people reflect on the collective, so that in societies where opportunities for contact between people of different social classes can be detrimental to social integration. It has been argued that the social distance among groups widened by extreme inequality can be described as an _empathy gulf_ that comprises barriers to imagining others’ lifestyles and goods, which can also be expressed in spatially segregated interactions and isolated lifeworlds that provide meaning to social inequality experiences, resulting in doubts about worthiness when the lower classes contrast themselves with extravagant lifestyles from the upper classes, undermining feelings of social inclusion and cohesion [@sachweh_moral_2012]. Consequently, segregation drives the lives of others to become more distant and might have consequences on empathy and solidarity toward others, potentially leading to perceiving fellow citizens as strangers [@otero_lives_2022, p. 758]. Thus, we suggest that attitudes towards economic inequality may be influenced by the class situation of both the individual and network ties, resembling the core claims of intergroup contact theory [@pettigrew_metaanalytic_2006] so that social influence processes can either align or polarize attitudes according to the characteristic of groups as well as their contact opportunities given the degree of segregation.

@@

In addition, studies that have more directly addressed the relationship between networks and social class have been consistent with previous findings. In this regard, evidence from Sweden suggests that people tend to assimilate their opinions according to their surrounding friends and acquaintances, which is reinforced by class homophily, where higher contacts in the managerial class negatively impact redistributive demands, while ties to the working class increase them arguing that “individuals take an impression from others and modify their attitudes accordingly” [@lindh_missing_2021, p. 698]. Similarly, additional socialization sources are cross-class ties through the family members. Family of origin is crucial for political socialization, where class interests and norms are nurtured in childhood and early adulthood. A study in the United States has shown that those tied to the upper class through parental relations are less supportive of redistribution and progressive taxation than those with working-class family backgrounds [@lee_consider_2023]. Besides, @paskov_crossclass_2022 argues that the class position of family and partners shapes preferences en tanto que los households share risk according to the class position of its members, where class homogeneity in ties polarizes preferences between lower and upper classes. In contrast, this gradient is blurred as heterogeneity increases according to the class position of the partner and family of origin.

In sum, the attitudes of individuals toward redistribution are influenced by their class-embedded social relations. Thus, the degree of class segregation in these networks plays a crucial role in shaping people's beliefs about income distribution and redistributive policies. Furthermore, the degree of contact between different classes and class homophily reinforces the assimilation of opinions according to surrounding friends and acquaintances. Therefore, we expect that being segregated in similar class environments polarizes own class interests, where segregated lower (upper) classes hold more (less) redistributive preferences.

@@

Income inequality can deteriorate social bonding and increase segregation.

* status anxiety among the upper-classes [@wilkinson_spirit_2010] and lower interpersonal trust  [@kragten_income_2017]

* limited opportunities for social mobility consequently reduce class permeability in social networks [@otero_differences_2023]

However, it has been discussed that inequality might also increase egalitarian preferences among those in better material situations.

* concerns about the negative consequences (e.g. crime) of inequality on overall social welfare, particularly in high-status groups (Altruism) [@rueda_who_2019]

>>>>

The class divide in political attitudes on the economic domain constitutes a core finding in policy attitudes, where material interests, normative motivations, and sociability practices intersect in attitude formation. Theoretically, the tradition in the social mobility field has defined classes as structural positions given their differences in labor market situation distinguished by employment relations, skills level, and authority within the workplace, which traditionally has focused on occupations instead of income or educational credentials [@Erikson1992;@Erikson1979]. In addition, political sociologists have systematically demonstrated that classes trace differences in other domains of social life as well, including ideological preferences [@evans_decline_1993], cultural consumption behavior [@Chan2007], and attitudes toward redistribution and welfare [@Svallfors2006]. Despite the claims of a supposed displacement of the distributive struggle between classes as a result of the salience of other group identities [@clark_declining_1993;@inglehart_culture_1990], during the last decades, the research has extensively developed on how class positions shape economic attitudes in the domains of redistribution, social security, and the welfare state [@lindh_class_2020, p. 421]. Empirically, the claimed blurring of the class cleavage has not been strongly supported, where class divides is clear regarding attitudes in the economic domain in general, such as perceived class conflicts [@perez_class_2023;@edlund_democratic_2015], attitudes toward market-based inequality [@andersen_preferences_2018;@lindh_public_2015], and redistributive preferences in particular [@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010].

In this regard, class-based explanations of redistributive attitudes have been mainly focused on the individual situation. Hence, one main claim is that economic preferences can be explained through the labor market situation, which comprises access to economic resources and exposure to unemployment risk [@meltzer_rational_1981;@rehm_risks_2009]. However, as material interests might prevail in scarcity, normative or value-driven motivations can also explain egalitarian economic preferences [@Kulin2013;@feldman_humanitarian_2001]. Besides, values become salient under greater security conditions but weaken under greater precariousness [@Maldonadoetal2019]. Alternatively, part of the literature suggests that the class divide can also be explained on top of the social interactions experienced in work relations [@oesch_redrawing_2006]. Given that workers spend a substantial part of their time at workplaces, social relations in the workspace might drive greater contact with diverse people in the case of interpersonal services, or vertical monitoring in managerial occupations can also imprint normative views that ultimately shape political opinions [@oesch_electoral_2018]. However, it is challenging to disentangle if the class differences are undoubtedly explained through work experiences because it is feasible to argue that political preferences are the result of early-life internalization of social norms, which might be crystalized in career self-selection paths, which also might be influenced by the organizational hiring requirements for specific jobs [@kitschelt_occupations_2014].

@@

As social relations within the occupational structure serve as the foundation for class analysis, widening our examination to other dimensions of social life, such as family dynamics, the development of friendships, and diverse social activities, can significantly enrich the scope of class analysis, particularly concerning its relevance in political attitudes formation. In principle, social stratification scholars have traditionally considered _differentiation_ as the observable patterns of association given by the intersection of status and group membership, which can be understood as relational networks between positions [@blau_inequality_1977]. Empirically, homophily is one of the most significant findings in network research, understood as the higher likelihood of tie formation between people of similar characteristics [@mcpherson_birds_2001; p.416] around demographic characteristics that trace friendship and family ties such as age and gender, as well as the segregated social environments alongside ethnicity and socioeconomic status [@bargsted_pautas_2020; @plaza_juntos_2022,@lazarsfeld_friendship_1954]. In addition, psychosocial factors such as socialization preferences and group dynamics also play a role in segregation processes, considering that isolated social environments decrease the interchange of world views, reinforcing in-group selectivity and value similarity [@visser_attitudes_2004;@homans_human_1951]. Nonetheless, without denying the relevance of socialization preferences, segregation has been predominantly explained through social differentiation in social activities as foci of contact opportunities that ultimately consolidate social networks [@feld_focused_1981; @mcpherson_network_2019]. From this scope, we argue that class relations represent not only resource-based distinctions but also patterns of sociability and exchange observed in the differentiation of social ties.

How classes are formed also provides insights regarding their social relations and sociability practices. In sociology, classes are understood as the basis for action, given their labor market situation, that ultimately shapes their access to opportunities and resources [@weber_economy_1978]. According to habitus theory, classes incur a series of distinctive practices pivoted in cultural capital as the basis for mutual recognition [@bourdieu_distinction_1984]. From this approach, classes seek to improve their position through the intergenerational transmission of resources, where similarities in friendship and family choices play a crucial role in the reproduction privilege. Additionally, while symbolic resources reinforce status distinctions, access to exclusive social activities in the upper class increases segregation at the expense of excluding the lower classes [@bourdieu_reproduccion_1981]. Ultimately, classes consolidate their demographic and cultural identities by mobilizing resources and shared sociability practices, ensuring intergenerational reproduction of their structural positions [@goldthorpe_sobre_1992]. Hence, the socialization of common lifestyles and shared worldviews facilitates the consolidation of social classes, where social resources embedded in social networks follow a pattern of accumulation and exclusion that ultimately leads to social segregation in both the upper and the lower bounds of the social structure.

Previous research on social class and tie formation has suggested that class permeability in social networks, understood as the formation of cross-class social ties, is traced differently across property and authority boundaries. Empirically, evidence from Western industrialized societies has shown that the formation of friendship ties between owners and the working class is significantly less likely, in contrast to the tendency of higher tie formation between supervisors and workers, suggesting that the property dimension is much less permeable than authority-based boundaries given that class interests increase the social distance between proprietors and workers, while the intermediate class position of supervisors and contact frequency with workers make friendship tie formation more likely [@wright_relative_1992]. Similarly, evidence from Chile has shown that the higher permeability of the intermediate classes contrasts with the more homogeneous networks of the working class, suggesting that their limited resources and lower capacity to be socially engaged ultimately result in a lack of social resources that leads to social isolation, while the upper class is less permeable and homogenous as a result of its tendency to self-selection as a practice that ultimately seeks to reproduce their privileged status positions [@otero_open_2021].

In contrast, cross-class embeddedness can be described regarding social activities and changes across generations and life course. In this regard, @pichler_social_2009 suggest that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the working class, which tends to be more homogeneous in its civic engagement behavior. Similarly, other studies in Europe and South America have exhibited that the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes, where being socially mobile does help to improve both dimensions in tie formation [@cepic_how_2020;@carrascosa_class_2023]. Nevertheless, it does not equalize the weight of class background compared to those intergenerationally stable in the upper class. Additionally, the few longitudinal evidence suggests that networks change following a dynamic of cumulative advantages regarding its composition in terms of prestige and diversity, where upper classes improve in both dimensions, while the lower class shows more stability across the life course [@volker_social_2020].

Altogether, these studies account for class-based network segregation as the degree of connectedness of an individual to different occupations is meant to represent social resources embedded in social networks vertically in the social structure [@lin_social_2007]. Drawing on this approach, the attempts of stratification scholars have been focused on how social connections are distributed across the class structure as a matter of social integration [@blau_macrosociological_1977]. Besides, _homogeneity_  is described as lacking cross-class network ties and is conceptually more proximate to the homophily principle because it is anchored in individual class positions. At the same time, _diversity_ is defined as the rate of dissimilar ties or simply the total ties to certain groups that do not necessarily count with a reference position to describe the network composition. Hence, both approaches provide alternatives for studying social class from a network perspective that has been increasingly discussed in the stratification literature and the scholarly discussion about social class and political attitudes.

@@

  Differently, it has been suggested that the lens of a network approach provides a better picture of the class relations that nurture social norms and group identity [@kalmijn_social_2007, p. 550]. This claim resembles the fact that classes are characterized as collectivities with differences in their degrees of cohesion and solidarity, encompassing unequal status-based social interactions that are linked to the individual or household material well-being, cultural perspectives, and political preferences that structure the broader social experiences [@morris_attenuation_1996, p. 48]. Furthermore, these social relations contribute to the degree to which people reflect on the collective, so that in societies where opportunities for contact between people of different social classes can be detrimental to social integration. It has been argued that the social distance among groups widened by extreme inequality can be described as an _empathy gulf_ that comprises barriers to imagining others’ lifestyles and goods, which can also be expressed in spatially segregated interactions and isolated lifeworlds that provide meaning to social inequality experiences, resulting in doubts about worthiness when the lower classes contrast themselves with extravagant lifestyles from the upper classes, undermining feelings of social inclusion and cohesion [@sachweh_moral_2012]. Consequently, segregation drives the lives of others to become more distant and might have consequences on empathy and solidarity toward others, potentially leading to perceiving fellow citizens as strangers [@otero_lives_2022, p. 758]. Thus, in a similar vein, as the intergroup contact approach has established [@pettigrew_metaanalytic_2006], we suggest that attitudes towards redistribution may be influenced by the class situation of both the individual and network ties where social influence processes can either align or polarize attitudes according to the characteristics of groups and the level of segregation that allows factual contact opportunities.

@@

In addition, studies that have more directly addressed the relationship between networks and social class have been consistent with previous findings. For instance, evidence from Sweden suggests that people tend to assimilate their opinions according to their surrounding friends and acquaintances, which is reinforced by class homophily, where higher contacts in the managerial class negatively influence support for redistribution, while ties to the working class increase it regardless of the individual class position suggesting that “individuals take an impression from others and modify their attitudes accordingly” [@lindh_missing_2021, p. 698]. Similarly, additional socialization sources are cross-class ties through the family members. Family of origin is crucial for political socialization, where class interests and norms are nurtured in childhood and early adulthood. A study in the United States has shown that those tied to the upper class through parental relations are less supportive of redistribution and progressive taxation than those with working-class family backgrounds [@lee_consider_2023]. Besides, @paskov_crossclass_2022 argues that the class position of family and partners shapes preferences while households share risk according to the class position of its members, where class homogeneity in ties polarizes preferences between lower and upper classes. In contrast, this gradient is blurred as heterogeneity increases according to the class position of the partner and family of origin.

In sum, we argue that redistributive preferences are influenced by their class-embedded social relations. Therefore, the degree of class segregation in these networks plays a crucial role in shaping support for redistribution. Furthermore, the degree of contact between different classes and class homophily reinforces the assimilation of opinions according to surrounding friends and acquaintances. Thus, we hypothesize that segregation in similar class environments polarizes own class interests, where segregated lower (upper) classes hold more (less) redistributive preferences.

@@

###  Consequences of Economic Inequality on Social Networks  
The consequences of economic inequality are diverse. Regarding social networks, the consequences of interpersonal trust, civic participation, and social resources have been previously reported in the literature. In addition, the unequal distribution of economic resources impacts coexistence through the socio-spatial segregation of the inhabitants of a territory.

### Consequences of Economic Inequality on redistributive preferences
* concerns about the negative consequences (e.g. crime) of inequality on overall social welfare, particularly in high-status groups (Altruism) [@rueda_who_2019]
